MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-19 main/MSstatsPTM_Main.tex
moved sections to main

The technological challenges in PTM identification and quantification challenge the downstream statistical analyses. Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as t-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~ \cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
}
\label{tab:dataDescription}
\end{table}

@@

The simulation provides an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. 1000 proteins were generated per dataset with 10 unmodified features per protein. One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

\todo{generated (how many?)  repeated synthetic datasets?}. \todo{I have a hard time following this section. Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein? Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from. It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

The custom spike-in experiment had 4 conditions with two biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

@@

The spike-in peptides were mixed with human lysate to create four mixture conditions. The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.  Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only. \todo{How the biological replicates in each condition were obtained?}

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
\todo{In this experiment XXX biological replicates were allocated to 1 TMT mixture in an unbalanced design.} Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified in 11-plex \todo{?} isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. \todo{One MS run total?}

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

@@

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced design. \todo{Further details in Suppler Sec XXX?}

@@

\todo{Would it be possible to re-organize this section? (1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation? Ok if not; (3) Existing methods: MSstats. Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure. I feel that the experimental design in Suppler is redundant. Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }

@@

Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions.  The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another \todo{add references that substantiate that claim}. Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approach can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

More recently, Isobar-PTM was proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows to experiments with complex designs.

Mathematical details regarding these existing methods is in Supplementary Sec. \ref{supp-sec:ttest}. \todo{Does the section describe Isobar-PTM as well?}

\subsection*{Proposed approach}

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Existing work in statistical modeling and parameter estimation}

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance \cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site \cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

Additionally, the proposed method can include imputing missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model \cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Detection of changes in PTMs - Balanced design with one source of variation}
\label{sec:adjustment_summary}

Here we propose a statistical method for detecting changes in PTMs and removing confounding with changes in global protein abundance. The method is shown when applied to a balanced design with multiple conditions and one technical replicate. In practice the experimental design can be unbalanced and variation can come from multiple sources. We show how the methods can be applied to more complex experimental designs in the next section.

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$. The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide. For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:

@@

Finally the degrees of freedom are calculated using the following formula.

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}.

@@

The statistical modeling approaches discussed above can also be extended to complex designs including experiments with additional sources of variation and unbalanced designs. Equation 1 is the same in this case and the null hypothesis is still that there is no difference in mean PTM abundance between Conditions $i$ and $i'$ after adjusting for changes in mean unmodified peptide abundance between Conditions $i$ and $i'$.

@@

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^z\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

@@

\todo{I do not think you need to repeat all the MassIVE.quant ids - just refer to the table in the beginning. Then add that in addition to the data the repository containers also include analysis scripts and results}

>>>>

The technological challenges in PTM identification and quantification challenge the downstream statistical analyses. Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
}
\label{tab:dataDescription}
\end{table}

@@

The simulation provides an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. 1000 proteins were generated per dataset with 10 unmodified features per protein. One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. The fold change of half the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$. The values for $\sigma^2$ were motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that? Devon: The simulation did not include any missing values due to low abundance. Should we mention this or just leave it as is?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

In the custom spike-in experiment 8 biological replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

@@

The spike-in peptides were mixed with human lysate to create four mixture conditions. The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.  Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
In this experiment 11 biological replicates were allocated to 1 TMT mixture in an unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. \todo{One MS run total? Only one MS run was used. Not sure the best way to say this.}

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run, with the same design, to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

@@

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run with the same design to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced design. Further details in Suppler Sec \ref{supp-sec:usp30}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Background Information}

\subsection*{Experimental Design}

\todo{Would it be possible to re-organize this section? (1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and $t$-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation? Ok if not; (3) Existing methods: MSstats. Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure. I feel that the experimental design in Suppler is redundant. Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }

\todo{Devon: I moved most everything around as recommended. The notation is mainly described once we start getting into the equations (t-test section). We can move this into the design section but I thought it was clearer to describe it with the equations. The results section now starts with the proposed method in the same notation. The supplementary now only includes the table and extra results.}

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. Additionally, some spectral features can be missing. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition $i$ is denoted by $i^{\ast}$.  The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Features corresponding to unmodified peptides are considered together, where the protein abundance in Condition $i$ is denoted by $i$. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed.

@@

Many investigations performed differential abundance analysis of PTMs using two-sample $t$-test or its extensions.  The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.

An overview of the existing approach using $t$-test and Limma are shown in the following sections.

\subsubsection*{Two-sample $t$-test}

Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]
Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]
where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\subsubsection*{Limma}

Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions. It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances. Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.

With respect to PTM analysis, feature level summarization and adjustment for changes in the unmodified peptides are performed in the same way as $t$-test. After fitting a linear model, the final variance is moderated using the global variance across all modified peptides. This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling PTM $i$ and $v_\Theta$ is the global variance.

\subsection*{Existing work in statistical modeling and parameter estimation}

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

Additionally, $MSstats$ provides a method for imputing missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model~\cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

\subsubsection*{Detection of changes in PTMs - Balanced design with one source of variation}
\label{sec:adjustment_summary}

We propose a statistical method for detecting changes in PTMs while removing confounding with changes in global protein abundance. Here the method is applied to a balanced design with multiple conditions and one technical replicate. In practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$ respectively. The feature and run level summarization for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide. For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:

@@

Finally the degrees of freedom are calculated as follows.

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the $t$ distribution with degrees of freedom approximated by the Satterthwaite method~\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$~\cite{Benjamini:1995}.

@@

The statistical modeling approaches discussed above can be extended to complex designs including experiments with additional sources of variation and unbalanced designs. In this case our null hypothesis and Equation 1 are unchanged.

@@

The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ and degrees of freedom in Equations 3 and 4 are unchanged from the balanced case, although $\hat{\sigma}_{\gamma^\ast}$ and $\hat{\sigma}_{\gamma}$ are now estimated using RML.

@@

The details of the four experiments, including raw data, and intermediate data processing outputs are located in individual MassIVE repositories, referenced in \ref{tab:dataDescription}.
